---
title: "PSTAT175 -- Project"
author: "Valeria Lopez, Ciaran Finnegan, Kartik Adimulam"
date: "June 2, 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(survival)
library(survminer)
library(ggplot2) 
library(tidyr)
library(xfun)
```

# 1. Introduction
## Overview of the Dataset

We are studying the Chemotherapy for Stage B/C `colon` dataset from the `survival` package in R (Laurie et al., 1989). This dataset comes from one of the first successful trials of chemotherapy for colon cancer. It examines the impact of treating patients with just a compound called Levamisole or a combination of it and 5-FU, a moderately toxic chemotherapy agent, or neither. Each patient has two records, one for recurrence and one for death, which we will take into account in our analysis.

Our dataset contains survival time data of days until death of colon cancer patients that received different levels of chemotherapy. The primary covariate in our analysis is the `rx` treatment variable which has 3 levels (no treatment, low-toxicity treatment, and moderate toxicity treatment) and the censoring status indicator. There are other covariates such as:  
  - `sex`: male (1) or female (0) \
  - `age`: in years \
  - `obstruct`: was the colon obstructed by the tumor? \
  - `perfor`: was the colon perforated? \
  - `adhere`: is the tumor on nearby organs? \
  - `nodes`: how many lymph nodes have cancer? \
  - `differ`: is the tumor differentiated? \
  - `extent`: how much has the tumor spread locally? \
  - `surg`: how long from surgery to registration? \
  - `node4`: indicator variable of more than 4 positive lymph nodes \
  - `etype`: did subject's cancer recur or death \
(Moertel et al., 1991)

In our analysis we will use a significance level of $\alpha=0.05$.
  
## Primary Scientific Question
 
Does Levamisole + 5FU chemotherapy treatment have a significant effect on time until colon cancer recurrence or death due to colon cancer?

## Subquestions

1. Does adding 5FU chemotherapy to the Levamisole treatment increase the effectiveness of treatment?
2. Does Levamisole + 5FU treatment affect time until recurrence differently than time until death?
3. Are there other covariates that significantly affect time until recurrence or death?

```{r, include = FALSE}
head(colon)
```

## Survival Probability of Treatments 

In our first step of evaluating which treatments affect survival times, we create a Kaplan-Meier estimate plot to visualize and gain a base understanding of how the 3 treatments in the dataset affect the survival probabilities of colon cancer patients. 

```{r, warning = FALSE}
colon=colon
colon_fit = survfit(Surv(time, status) ~ rx, colon)
ggsurvplot(colon_fit, colon, title='KM Estimator Colon Cancer',xlab='time',
           ylab='Survival Prob.',surv.median.line = 'hv',palette = "Dark2", 
           conf.int = TRUE,    
           risk.table = TRUE,
           ggtheme = theme_minimal(), 
           font.main = c(16, "bold", "black"),  
           font.x = c(14, "bold"),            
           font.y = c(14, "bold"))
```
From the graph, we observe that the patients with the treatment that combines 5FU and Levamisole have a higher survival probability than those with no treatment and those with just the Levamisole treatment. The survival probability for patients with the Levamisole treatment seem to not have a large difference compared to those with no treatment, but the patients with no treatment have the lowest survival time. 

## Failure type and Models explanation

An interesting and important feature of our data set is that there are two different types of events for each subject, recurrence and death. This poses a potential problem for us because these two event types are very different in nature. In addition it is possible for one of them to be censored while the other is not. There are 2 rows for each subject in this data, each representing one of the event types. It is possible for a subject to have a recurrence at time 1000, for example, but then be censored at type 1500. In this case, the data set displays a status=1 for the recurrence row and a status=0 for the death row. Since our overall goal in this analysis is to understand the effectiveness of the Levamisole+5FU chemotherapy treatment, we will employ some different methods in order to understand the effect on time (both recurrence and death). 

As advised, we will first split up our data based on event type, creating two coxPH models. One of the models will be modelling the effect of Lev+5FU treatment on time until recurrence. The other model will be modelling the effect of Lev+5FU treatment on time until death. These models will be completely independent of each other and will proceed in the normal Cox PH fashion, stratifying on variables that are in violation of the PH assumption and using AIC/BIC scores to evaluate covariates added into the model. 

However, it is important also to not only study both failure types in isolation, but to combine them into a single model. We will do this in a myriad of ways. The first technique that we will employ is to use the entire data set, but then to simply stratify on event type. In this case, we will create a single model that estimates a different baseline hazard for recurrence and death. For each of these baseline hazards, we will attempt to estimate the overall effect of Lev+5FU on time. The coefficient estimate that we derive will be a single value that estiamtes the equal effect of Lev+5FU on each event type.

Another technique that we will employ is the Cause-Specific Cox PH Hazard Model. In this model, we treat other event types as censored, allowing us to determine a separate hazard rate for each event. This does not give us an overall effect of Lev+5FU on time, which is our goal. Instead we want to treat each failure type differently with a different coefficient estimate. This appraoch is slightly similar in nature to splitting up our data set, but also different in the treatment of censoring. With Cause-Specific Cox PH, ALL OTHER event types are treated as censored. 

## Recurrence Analysis

```{r}
colon.recur = colon[colon$etype==1,]
colon.death = colon[colon$etype==2,]

colon.recur.cox = coxph(Surv(time,status)~rx, colon.recur)
summary(colon.recur.cox)
anova(colon.recur.cox)


```

In our time until recurrence model, treatment using Levamisole compared with the baseline effect has a $p$-value of 0.888. This teaches us that this low-toxicity compound is not effective in reducing the time until recurrence for colon cancer patients.

However, the $p$-value of the treatment which combines 5FU (moderately toxic) treatment and Levamisole (low toxicity) is 0.00002. We can conclude that, at the given significance level, the combination of 5FU and Levamisole is effective in reducing time until recurrence but the treatment of just using Levamisole is not. 

Overall, we observe using a likelihood ratio test $p$-value of ~0 that the treatment itself is effective in reducing time until recurrence, but most of this effect is likely due to the high effectiveness of Levamisole+5FU treatment.


## Recurrence CI

```{r}
exp(confint(colon.recur.cox))

```

The CI show that Levamisole treatment is not significant but Levamisole+5FU treatment is effective in reducing time until recurrence. Our hazard ratio CI for Lev treatment contains 1 for both models, but our hazard ratio CI for Lev+5FU treatment does not contain 1 for the model. We are 95% confident that Lev+5FU treatment significantly affects time until recurrence.


## Recurrence Hazard Ratio

```{r}
exp(coef(colon.recur.cox))
```

At any point in time Lev+5FU treated patients are 0.6 times as likely to have a colon cancer recurrence compared to patients without treatment. 

\newpage

## Recurrence Model Fitting
Using forward stepwise selection  and AIC, we can find the most relevant covariates to build and finds the best model at predicting survival times. We first perform this process on the subset of data where there was a recurrence of cancer in patients. 
```{r, warning = FALSE}
colon.rx.sex = coxph(Surv(time,status)~rx + sex, colon.recur)
colon.rx.age = coxph(Surv(time,status)~rx + age, colon.recur)
colon.rx.obstruct = coxph(Surv(time,status)~rx + obstruct, colon.recur)
colon.rx.perfor = coxph(Surv(time,status)~rx + perfor, colon.recur)
colon.rx.adhere = coxph(Surv(time,status)~rx + adhere, colon.recur)
colon.rx.nodes = coxph(Surv(time,status)~rx + nodes, colon.recur)
colon.rx.differ = coxph(Surv(time,status)~rx + differ, colon.recur)
colon.rx.extent = coxph(Surv(time,status)~rx + extent, colon.recur)
colon.rx.surg = coxph(Surv(time,status)~rx + surg, colon.recur)
colon.rx.node4 = coxph(Surv(time,status)~rx + node4, colon.recur)

AIC(colon.rx.sex, colon.rx.age, colon.rx.obstruct, colon.rx.perfor,
    colon.rx.adhere, colon.rx.nodes, colon.rx.differ, colon.rx.extent,
    colon.rx.surg, colon.rx.node4)
```

The best model is adding nodes as a covariate, attaining an AIC of 5827.4. The second best model is differentiation of tumor and the third best is node4 (more than 4 positive lymph nodes). However, we should not include node4 since it is highly correlated with nodes. They are essentially measuring the same thing and in the dataset description it says it is unknown which of the two variables is correct, so this is a somewhat arbitrary selection.

```{r,  warning = FALSE}
colon.rx.nodes.differ = coxph(Surv(time,status)~rx + nodes + differ, colon.recur)
colon.rx.nodes.sex = coxph(Surv(time,status)~rx + nodes + sex, colon.recur)
colon.rx.nodes.age = coxph(Surv(time,status)~rx + nodes + age, colon.recur)
colon.rx.nodes.obstruct = coxph(Surv(time,status)~rx + nodes + obstruct, colon.recur)
colon.rx.nodes.perfor = coxph(Surv(time,status)~rx + nodes + perfor, colon.recur)
colon.rx.nodes.adhere = coxph(Surv(time,status)~rx + nodes + adhere, colon.recur)
colon.rx.nodes.extent = coxph(Surv(time,status)~rx + nodes + extent, colon.recur)
colon.rx.nodes.surg = coxph(Surv(time,status)~rx + nodes + surg, colon.recur)

AIC(colon.rx.nodes.differ, colon.rx.nodes.sex, colon.rx.nodes.age, colon.rx.nodes.obstruct, colon.rx.nodes.perfor, colon.rx.nodes.adhere, colon.rx.nodes.extent, colon.rx.nodes.surg)
```
Adding differ as the next covariate is the best model, as our model now has an AIC of 5676.21 

```{r}
colon.rnd.sex = coxph(Surv(time,status)~rx + nodes + differ + sex, colon.recur)
colon.rnd.age  = coxph(Surv(time,status)~rx + nodes + differ + age, colon.recur)
colon.rnd.obstruct  = coxph(Surv(time,status)~rx + nodes + differ + obstruct, colon.recur)
colon.rnd.perfor  = coxph(Surv(time,status)~rx + nodes + differ +perfor , colon.recur)
colon.rnd.adhere = coxph(Surv(time,status)~rx + nodes + differ +adhere , colon.recur)
colon.rnd.extent = coxph(Surv(time,status)~rx + nodes + differ + extent, colon.recur)
colon.rnd.surg  = coxph(Surv(time,status)~rx + nodes + differ +surg , colon.recur)
AIC(colon.rnd.sex, colon.rnd.age, colon.rnd.obstruct, colon.rnd.perfor, colon.rnd.adhere, colon.rnd.extent, colon.rnd.surg)
```
Adding extent as the next covariate is the best model, attaining an AIC of 5658.7

```{r,  warning = FALSE}
BIC(colon.rnd.extent,colon.rx.nodes.differ, colon.rx.nodes, colon.recur.cox)
```
Using BIC as a sanity check helps us to verify that adding extent as another covariate is still beneficial to our model's prediction accuracy while avoiding overfitting (too many covariates), attaining a BIC of 5679.21. 

```{r}
colon.rnde.sex = coxph(Surv(time,status)~rx + nodes + differ + extent + sex, colon.recur)
colon.rnde.age = coxph(Surv(time,status)~rx + nodes + differ + extent + age, colon.recur)
colon.rnde.obstruct = coxph(Surv(time,status)~rx + nodes + differ + extent + obstruct, colon.recur)
colon.rnde.perfor = coxph(Surv(time,status)~rx + nodes + differ + extent + perfor, colon.recur)
colon.rnde.adhere = coxph(Surv(time,status)~rx + nodes + differ + extent + adhere, colon.recur)
colon.rnde.surg = coxph(Surv(time,status)~rx + nodes + differ + extent + surg, colon.recur)

AIC(colon.rnde.sex, colon.rnde.age, colon.rnde.obstruct, colon.rnde.perfor,
    colon.rnde.adhere, colon.rnde.surg)
BIC(colon.rnde.sex, colon.rnde.age, colon.rnde.obstruct, colon.rnde.perfor,
    colon.rnde.adhere, colon.rnde.surg)
```
Adding Surg (time from surgery until registration into the study) as our next covariate again lowers our AIC to 5655.55

```{r}
colon.rndes.sex = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + sex, colon.recur)
colon.rndes.age = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + age, colon.recur)
colon.rndes.obstruct = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + obstruct, colon.recur)
colon.rndes.perfor = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + perfor, colon.recur)
colon.rndes.adhere = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + adhere, colon.recur)

AIC(colon.rndes.sex, colon.rndes.age, colon.rndes.obstruct, colon.rndes.perfor, colon.rndes.adhere)
```
We attain a slightly lower AIC of 5654.37 by adding sex as our next covariate. However, since the change is small we will verify this using BIC score too. 

```{r}
BIC(colon.rndes.sex, colon.rnde.surg)
```
Using BIC score instead, which penalizes more covariates more intensely than AIC, we see that our previous model is actually preferable, so we will stop adding covariates at Surg. 

```{r}
anova(colon.rndes.sex)
```
We also verify this by checking that if we add sex as a covariate, it is not significantly affecting time until recurrence from the likelihood ratio test.


### Checking PH Assumptions of Recurrence

```{r}
colon.fit = survfit(Surv(time, status) ~ rx, colon.recur)
ggsurvplot(colon.fit, colon.recur, 
           legend.labs=c('Level 1','Level 2', 'Level 3'), 
           fun='cloglog') + 
  labs(x='Days until Recurrence')

cox.zph(colon.recur.cox)
```
The proportional hazards assumption seems violated from the log-log plot for Baseline and low toxicity treatment, but we have already observed that those two treatments do not significantly differ in hazard rates. It appears that the difference between low toxicity and medium toxicity does not violate the proportional hazards assumption, and we verify this using the Schoenfield residuals test attaining a p-value of 0.86 for the rx covariate. 

```{r}
cox.zph(colon.rnde.surg)
colon.fit.differ = survfit(Surv(time, status) ~ differ, colon.recur)
ggsurvplot(colon.fit.differ, colon.recur, 
           legend.labs=c('Well','Moderate', 'Poor'), fun='cloglog') +
  labs(x='Days until Recurrence')
```
From our cox ZPH test we see that the differentiation of tumor covariate violates the proportional hazards assumption with a p-value of 0.00024 and we verify this using a log-log plot. We can observe many cross-over points between the plot for 'moderate' differentiation and 'well' differentiation. To account for this violation, we will attempt to stratify on the differentiation covariate and produce a stratified Cox PH model. 

```{r}
colon.rnde.surg = coxph(Surv(time,status)~rx + nodes + strata(differ) + extent + surg, colon.recur)
anova(colon.rnde.surg)
cox.zph(colon.rnde.surg)

mean(colon.recur$nodes, na.rm=TRUE)
max(colon.recur$nodes, na.rm=TRUE)
```
Our AIC drops to 4967.46 after stratifying on the differentiation variable instead of simply including it as a covariate in our model. This is a significant decrease in AIC most likely due to the fact that differentiation should have never been included as a covariate in the model, as it violated the proportional hazards assumption. 

To verify again that none of our covariates violate this assumption, we will run the Cox ZPH test on our new stratified model. We observe that all of our p-values are greater than 0.05 and we can keep them in our model as covariates. 

We verify again post-stratification that all of our covariates are significant. We can begin our model interpretation as follows:

- After accounting for other covariate effects, the impact on hazard rate of the Levamisole+5FU treatment remains a factor of 0.6.  
- For each additional lymph nodes with detectable cancer that is observed at the start of the study, the hazard rate for that individual increases by a factor of 1.078. This is not a huge effect, but it is important to note that the number of lymph nodes with cancer were as high as 33 in our data set, so the hazard rate could gradually increase to significantly high levels as this covariate increases for certain subjects. However, the mean of this covariate is only ~ 3.7 nodes
- For our 4 levels of increasing local spread extent of the colon cancer, hazard rate increases by an alarming factor of 1.66. This is an intuitive finding, as the extent of cancer spread is a strong indicator of a patient having a recurrence or death. 
- If time from surgery to registration is classified as 'long', hazard rate is increased by a factor of 1.28. This is an interesting finding and indicates that the time spent between surgery and registration is vital to the time it took patients to have a disease recurrence or death. This is also an intuitive finding, as a patient that had a long time in between is experiencing more time that could contribute to the growth of cancer spread. 

## Death Analysis

Using the same exact process as above for the data with etype=1 (recurrence of cancer), we find that for etype = 2 (death), the model had the same covariates until adding the 6th covariate. Adding obstruct to the death cox model gave us the next best model which is different from the recur data, where at this point, adding sex as our next covariate gave us the lower AIC score. However, in both subsets of data, adding the 6th covariate gave us a higher BIC score, allowing us to leave it out and have the same covariates (rx, node, differ, extent, surg).

```{r}
colon.death.cox = coxph(Surv(time, status) ~ rx, colon.death)
summary(colon.death.cox)
anova(colon.death.cox)
```

In our time until death model, treatment using Levamisole compared with the baseline effect has a $p$-value of 0.809. This teaches us that this low-toxicity compound is not effective in reducing the time until death for colon cancer patients.

However, the $p$-value of the treatment which combines 5FU (moderately toxic) treatment and Levamisole (low toxicitity) is 0.0018. We can conclude that the combination of 5FU and Levamisole is effective in reducing time until death but the treatment of just using Levamisole is not. This is an outcome that is discussed in the original study, which states that "after correction for imbalances in prognostic variables, were only suggestive for levamisole alone (P = .05) but quite significant for levamisole plus 5-FU (P = .003)" (Laurie et al., 1989). It is interesting, however, that our p-value for levamisole treatment alone obtained from our cox PH model is so much higher than the p-value obtained by the researchers of this study. Although both of us agree that Levamisole treatment in itself in ineffective, our level of ineffectiveness is much higher than that of the study. This could potentially be attributed to the treatment of other covariates by the researchers differing from our own treatment. 

Overall, we observe using a likelihood ratio test p-value of 0.0023 that the treatment itself is effective in reducing time until death, but most of this effect is likely due to the high effectiveness of Levamisole+5FU treatment. 

##Death CI

```{r}
exp(confint(colon.death.cox))
```

The CI show that Levamisole treatment is not significant but Levamisole+5FU treatment is effective in reducing time until death. Our hazard ratio CI for Lev treatment contains 1 for the model, but our hazard ratio CI for Lev+5FU treatment does not contain 1. We are 95% confident that Lev+5FU treatment significantly affects time until death.

```{r}
exp(coef(colon.death.cox))
```

The patients with Lev+5FU are 0.69 times as likely to have a colon cancer death compared to patients without treatment. 


```{r, include = FALSE}
death.rx.sex = coxph(Surv(time,status)~rx + sex, colon.death)
death.rx.age = coxph(Surv(time,status)~rx + age, colon.death)
death.rx.obstruct = coxph(Surv(time,status)~rx + obstruct, colon.death)
death.rx.perfor = coxph(Surv(time,status)~rx + perfor, colon.death)
death.rx.adhere = coxph(Surv(time,status)~rx + adhere, colon.death)
death.rx.nodes = coxph(Surv(time,status)~rx + nodes, colon.death)
death.rx.differ = coxph(Surv(time,status)~rx + differ, colon.death)
death.rx.extent = coxph(Surv(time,status)~rx + extent, colon.death)
death.rx.surg = coxph(Surv(time,status)~rx + surg, colon.death)
death.rx.node4 = coxph(Surv(time,status)~rx + node4, colon.death)

AIC(death.rx.sex, death.rx.age, death.rx.obstruct, death.rx.perfor,
    death.rx.adhere, death.rx.nodes, death.rx.differ, death.rx.extent,
    death.rx.surg, death.rx.node4)

```

The best model in the data for patients with a recorded death adds nodes with an AIC of 5620.639. The second best model adds differ with an AIC of 5678.121 and the third best model adds node4 with an AIC of 5764.921. 

```{r, include = FALSE}
death.rx.nodes.differ = coxph(Surv(time,status)~rx + nodes + differ, colon.death)
death.rx.nodes.sex = coxph(Surv(time,status)~rx + nodes + sex, colon.death)
death.rx.nodes.age = coxph(Surv(time,status)~rx + nodes + age, colon.death)
death.rx.nodes.obstruct = coxph(Surv(time,status)~rx + nodes + obstruct, colon.death)
death.rx.nodes.perfor = coxph(Surv(time,status)~rx + nodes + perfor, colon.death)
death.rx.nodes.adhere = coxph(Surv(time,status)~rx + nodes + adhere, colon.death)
death.rx.nodes.extent = coxph(Surv(time,status)~rx + nodes + extent, colon.death)
death.rx.nodes.surg = coxph(Surv(time,status)~rx + nodes + surg, colon.death)

AIC(death.rx.nodes.differ, death.rx.nodes.sex, death.rx.nodes.age, death.rx.nodes.obstruct, death.rx.nodes.perfor, death.rx.nodes.adhere, death.rx.nodes.extent, death.rx.nodes.surg)

```

After adding differ we got an AIC of 5458.958.

```{r, include = FALSE}
death.rnd.sex = coxph(Surv(time,status)~rx + nodes + differ + sex, colon.death)
death.rnd.age  = coxph(Surv(time,status)~rx + nodes + differ + age, colon.death)
death.rnd.obstruct  = coxph(Surv(time,status)~rx + nodes + differ + obstruct, colon.death)
death.rnd.perfor  = coxph(Surv(time,status)~rx + nodes + differ +perfor , colon.death)
death.rnd.adhere = coxph(Surv(time,status)~rx + nodes + differ +adhere , colon.death)
death.rnd.extent = coxph(Surv(time,status)~rx + nodes + differ + extent, colon.death)
death.rnd.surg  = coxph(Surv(time,status)~rx + nodes + differ +surg , colon.death)

AIC(death.rnd.sex, death.rnd.age, death.rnd.obstruct, death.rnd.perfor, death.rnd.adhere, death.rnd.extent, death.rnd.surg)

```

Adding extent give us an AIC of 5441.828.

```{r, include = FALSE}
death.rnde.sex = coxph(Surv(time,status)~rx + nodes + differ + extent + sex, colon.death)
death.rnde.age = coxph(Surv(time,status)~rx + nodes + differ + extent + age, colon.death)
death.rnde.obstruct = coxph(Surv(time,status)~rx + nodes + differ + extent + obstruct, colon.death)
death.rnde.perfor = coxph(Surv(time,status)~rx + nodes + differ + extent + perfor, colon.death)
death.rnde.adhere = coxph(Surv(time,status)~rx + nodes + differ + extent + adhere, colon.death)
death.rnde.surg = coxph(Surv(time,status)~rx + nodes + differ + extent + surg, colon.death)

AIC(death.rnde.sex, death.rnde.age, death.rnde.obstruct, death.rnde.perfor,
    death.rnde.adhere, death.rnde.surg)


```

Adding surg gives an AIC of 5438.274.

```{r, include = FALSE}
death.rndes.sex = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + sex, colon.death)
death.rndes.age = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + age, colon.death)
death.rndes.obstruct = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + obstruct, colon.death)
death.rndes.perfor = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + perfor, colon.death)
death.rndes.adhere = coxph(Surv(time,status)~rx + nodes + differ + extent + surg + adhere, colon.death)

AIC(death.rndes.sex, death.rndes.age, death.rndes.obstruct, death.rndes.perfor, death.rndes.adhere)
```

Adding obstruct gives us an AIC of 5436.907, which is different from the recur data, where at this point, adding sex as out next covariate gave us the lower AIC score. 

```{r, include = FALSE}
BIC(death.rnde.surg, death.rndes.obstruct)

```
By using BIC, which penalizes the number of covariates more than AIC, we can see that the previous model, without obstruct, is a better model as it gives a BIC of 5462.657. Therefore, we will use the exact same covariates as the recurrence model. 


### Checking PH Assumptions for Death 

```{r}
cox.zph(death.rnde.surg)

colon.fit.differ = survfit(Surv(time, status) ~ differ, colon.death)
ggsurvplot(colon.fit.differ, colon.death, legend.labs=c('Well','Moderate', 'Poor'), fun='cloglog') + 
  labs(x='Days until Recurrence')
```

We can also observe, that similarly to our recurrence model the differentiation covariate again violates the PH assumption. We can verify this using a log-log plot, which shows many cross-over points bewteen 'moderate' differentiation and 'well' differentiation. We will remedy this using a stratification on the differentiation variable. 

```{r}
death.rnde.surg <- coxph(Surv(time,status)~rx + nodes + strata(differ) + extent + surg, colon.death)
AIC(death.rnde.surg)
```

Our AIC drops to 4769.91 after stratifying on the differentiation covariate instead of simply using it as a covariate in our model. This is a significant decrease and proves that our previous model had an alarming violation of the PH assumption. 

```{r}
cox.zph(death.rnde.surg)
```

We run cox ZPH again on this new stratified cox PH model to ensure no violations and observe that we have successfully remedied our issue. All of our p-values from the cox ZPH test are above 0.05. 

```{r}
anova(death.rnde.surg)
exp(coef(death.rnde.surg))
```

We verify again post-stratification that all of our covariates are significant. We can begin our model interpretation as follows:

- After accounting for other covariate effects, the impact on hazard rate of the Levamisole+5FU treatment is slightly increased to a factor of 0.694. In other words, its influence on decreasing hazard rate is slightly lessened after accounting for other variables. 
- For each additional lymph nodes with detectable cancer that is observed at the start of the study, the hazard rate for that individual increases by a factor of 1.088
- For our 4 levels of increasing local spread extent of the colon cancer, hazard rate increases by an alarming factor of 1.66.
- If time from surgery to registration is classified as 'long', hazard rate is increased by a factor of 1.29.


\newpage

# 3. Advanced Models 

## Cause Specific Cox PH model

Since the nature of our two events are different, one being recurrence of colon cancer while the other is death due to colon cancer, we will also construct a competing risks model, in which we consider the difference between the two outcomes. As mentioned before, a cause-specific coxPH model treats other event types as censored. For example, if our subject has a recurrence at time 1000 and a death at time 1500. The time 1000 observations is treated as if another person in the risk set was censored. The time 1500 event is treated as an actual event 

```{r}
recurrence_curv = survfit(Surv(time, etype==1)~rx, colon)
ggsurvplot(recurrence_curv)

death_curv= survfit(Surv(time, etype==2)~rx, colon)
ggsurvplot(death_curv)
```


```{r}

cause.recur = coxph(Surv(time, etype==1) ~ rx, colon)
anova(cause.recur)
exp(confint(cause.recur))
exp(coef(cause.recur))
```

We can see that Levamisole + 5FU treatment still has a significant effect on time until recurrence while Levamisole treatment by itself does not. In our cause-specific model, Lev+5FU treatment decreases recurrence hazard rate by 22%

```{r}
cause.death = coxph(Surv(time, etype==2) ~ rx, colon)
anova(cause.death)
exp(confint(cause.death))
exp(coef(cause.death))
```

We can see that Levamisole + 5FU treatment still has a significant effect on time until death  while Levamisole treatment by itself does not. In our cause-specific model, Lev+5FU treatment decreases death hazard rate by 26%.

##Stratifying on Event Type and Trying Interactions

To observe the distributed effect of the treatment across different event types (recurrence or death), we will now construct our competing risks model using the stratification method. By stratifying on the outcome variable (etype), we can create different baseline hazard functions for each event type and then observe the effect of the treatment on both baseline hazard functions. Overall, we should be able to attain an overall effect of the treatment on each type of outcome. 

```{r}
all.cox = coxph(Surv(time, status) ~ rx + strata(etype), colon)
anova(all.cox)
exp(confint(all.cox))
exp(coef(all.cox))
```

After stratifying on event type, we observe with a p-value of 1.92 * 10^-8 and a LRT statistic of 35.54 that the treatment of Levamisole + 5FU has a significant effect on time until recurrence and death. Again, Levamisole treatment in itself is not significant, as we can see from its 95% CI that it contains 1. We see one of our largest effect of Lev+5FU treatment in this model, as it decreases hazard rate of recurrence and death by 36%


```{r}
exp(coef(all.cox)[2])
exp(confint(all.cox))
```
With a 95% CI of [0.842, 1.139] for Levamisole treatment and [0.544,0.757] for Levamisole+5FU treatment, we can conclude with 95% confidence that Levamisole treatment is only effective with the 5FU agent added. It has an effect of reducing the hazard rate by 36%.

## Different effect across Event Types

We will now attempt to analyze if the treatment effect is different across the different event types. Again, our two event types are either a recurrence of cancer or a death due to cancer. Using an interaction term in our coxPH model, we can observe if the effect of Levamisole+5FU treatment differs between recurrence and death. We will stratify on the etype variable because we do not want to use it as a covariate. 

```{r}
diff.cox = coxph(Surv(time, status) ~ rx*strata(etype), colon)
anova(diff.cox)
```

We can conclude with our likelihood ratio test statistic of 0.6204 that the effect of the treatment does not differ significantly across event types. Therefore, we do not need to consider the interaction between rx treatment and etype. 

## Event Type as a Covariate

In order to more thoroughly understand the effect of event type, we will create a model where we actually use etype as a covariate instead of just stratifying on it. This will allow us to understand if etype itself as an impact on overall survival time and if there is a significant interaction between the effect of etype and the treatment. 

```{r}
etype.cox = coxph(Surv(time, status) ~ rx*etype, colon)
anova(etype.cox)
cox.zph(etype.cox)
```

We see from this model that our interaction between the treatment and etype is not significant, This means that event type has a significant effect on time, but this is not an important finding for us in context of our overall scientific question. The interaction between etype and rx treatment is not significant. We can see that it is best to stratify on etype instead of actually using it as a covariate in an interaction like we have done here. We do not want to estimate a coefficient for etype, but instead understand how the treatment affects both event types differently.

##CIC Curves

We will continue our competing risks model below with a CIC curve:

```{r}
colon$nstatus=factor(colon$nstatus)
multi.ph = survfit(Surv(time, nstatus)~1,colon)
ggcompetingrisks(multi.ph) 
```

We can see that the risk of recurrence is higher than the risk of death over time. This is predictable from our data since it is possible for a patient to have a recurrence before their death but not the other way around. We will now construct this same plot but observe the different CIC curves across the rx treatment variable

```{r}
multi.ph.d = survfit(Surv(time, nstatus)~rx,colon)
ggcompetingrisks(multi.ph.d)
```

We observe a similar phenomenon as before across all treatment types. However, in the Levamisole+5FU treatment group, the risk difference between the death and recurrence group is more narrow than the others. 

## Recurrent Events CP Model

To produce a combined model that considers both recurrence and death, we will restructure our data into stop-start data format. We will use a stratified counting process model The first occurrence for each subject is a recurrence of cancer, while the second occurrence is death. This model, however, essentially treats both failure types as the same event. 

## What Does a Failure Mean in the CP Model?

In the recurrent event model structured as a counting process model, the first stop value refers to a recurrence of cancer in a particular subject while the second stop value refers to a death of a patient. We are modelling recurrence as an initial stop value and death as a final stop value. 

A failure in the CP model refers to either a recurrence of cancer or a death due to colon cancer. There are no situations where etype 1 is censored while etype 2 is an event. It is only possible in our data set that etype 1 occurs while etype 2 is censored. This would mean that a subject had a recurrence of cancer at time 1000, for example, and then at time 1500 was censored and dropped from the study. It is not possible the other way around, since if an individual is censored in their etype=1 row, their etype 2 is automatically censored at that same recorded time. 

```{r}
colon$start=0
colon$stop = colon$time

for(i in 1:1858){
  ifelse(colon$etype[i]==2, colon$start[i] <- colon$time[i+1], colon$stop[i]<-colon$time[i] )
}

for(i in 1:1858){
  if(colon$stop[i]==colon$start[i]){
    colon$stop[i] <- colon$start[i]+0.01
  }
}

colon.cox = coxph(Surv(start,stop,status)~rx, colon)
summary(colon.cox)
anova(colon.cox)

colon.CI = exp(confint(colon.cox))
colon.CI
```
We confirm with 95% confidence that Levamisole treatment is not effective in itself using a confidence interval [0.475, 1.14] that contains 1. Our confidence interval for Levamisole+5FU treatment is quite small [0.58,0.8] and does not contain 1. We confirm our previous conclusion with 95% confidence that the combined treatment is effective in reducing time until recurrence or death. 


```{r}
exp(coef(colon.cox))
```


Lev+5FU reduces the hazard rate by 32%. We conclude that the treatment is indeed effective in significantly reducing time until a recurrence or death. This is the same conclusion that was derived from the recurrence data set as well as the death data set, but we can confirm that the combination treatment of low-toxicity and medium-toxicity compounds retains its effectiveness in the combined model of both event types. 

## Adding Covariates to the Combined Data Set

Again using forward stepwise selection, AIC scores, and verifying AIC scores using BIC when improvements are minimal, we will add the other covariates into our full model.

```{r, include = FALSE}
full.rx.sex = coxph(Surv(start, stop,status)~rx + sex + strata(etype), colon)
full.rx.age = coxph(Surv(start, stop,status)~rx + age+ strata(etype), colon)
full.rx.obstruct = coxph(Surv(start, stop,status)~rx + obstruct+ strata(etype), colon)
full.rx.perfor = coxph(Surv(start, stop,status)~rx + perfor+ strata(etype), colon)
full.rx.adhere = coxph(Surv(start, stop,status)~rx + adhere+ strata(etype), colon)
full.rx.nodes = coxph(Surv(start, stop,status)~rx + nodes+ strata(etype), colon)
full.rx.differ = coxph(Surv(start, stop,status)~rx + differ+ strata(etype), colon)
full.rx.extent = coxph(Surv(start, stop,status)~rx + extent+ strata(etype), colon)
full.rx.surg = coxph(Surv(start, stop, status)~rx + surg+ strata(etype), colon)
full.rx.node4 = coxph(Surv(start, stop,status)~rx + node4+ strata(etype), colon)

AIC(full.rx.sex, full.rx.age, full.rx.obstruct, full.rx.perfor,
    full.rx.adhere, full.rx.nodes, full.rx.differ, full.rx.extent,
    full.rx.surg, full.rx.node4)
```
Similar to our data subsets where etype = 1 and etype = 2, adding nodes as the next covariate after rx gave us the lowest AIC, 9597.89. We will not use node4 as a covariate since it is essentially measuring the same quantity as the nodes covariate. We do not want to use highly correlated covariates in our cox PH model. 

```{r, include = FALSE}
full.rn.sex = coxph(Surv(start, stop, status) ~ rx + nodes + sex + strata(etype), colon)
full.rn.age = coxph(Surv(start, stop, status) ~ rx + nodes + age + strata(etype), colon)
full.rn.obstruct = coxph(Surv(start, stop, status) ~ rx + nodes + obstruct + strata(etype), colon)
full.rn.perfor = coxph(Surv(start, stop, status) ~ rx + nodes + perfor + strata(etype), colon)
full.rn.adhere = coxph(Surv(start, stop, status) ~ rx + nodes + adhere + strata(etype), colon)
full.rn.differ = coxph(Surv(start, stop, status) ~ rx + nodes + differ + strata(etype), colon)
full.rn.extent = coxph(Surv(start, stop, status) ~ rx + nodes + extent + strata(etype), colon)
full.rn.surg = coxph(Surv(start, stop, status) ~ rx + nodes + surg + strata(etype), colon)

AIC(full.rn.sex, full.rn.age, full.rn.obstruct, full.rn.perfor,
    full.rn.adhere, full.rn.differ, full.rn.extent,
    full.rn.surg)
```

Adding differ gives us the lowest AIC of 9329.79 

```{r, include = FALSE}
full.rnd.sex = coxph(Surv(start, stop, status) ~ rx + nodes + differ + sex + strata(etype), colon)
full.rnd.age = coxph(Surv(start, stop, status) ~ rx + nodes + differ + age + strata(etype), colon)
full.rnd.obstruct = coxph(Surv(start, stop, status) ~ rx + nodes + differ + obstruct + strata(etype), colon)
full.rnd.perfor = coxph(Surv(start, stop, status) ~ rx + nodes + differ + perfor + strata(etype), colon)
full.rnd.adhere = coxph(Surv(start, stop, status) ~ rx + nodes + differ + adhere + strata(etype), colon)
full.rnd.extent = coxph(Surv(start, stop, status) ~ rx + nodes + differ + extent + strata(etype), colon)
full.rnd.surg = coxph(Surv(start, stop, status) ~ rx + nodes + differ + surg + strata(etype), colon)


AIC(full.rnd.sex, full.rnd.age, full.rnd.obstruct, full.rnd.perfor,
    full.rnd.adhere, full.rnd.extent,
    full.rnd.surg)
```
Adding extent as our next covariate gives us an AIC of 9313.93

```{r, include = FALSE}
full.rnde.sex = coxph(Surv(start, stop, status) ~ rx + nodes + differ + extent + sex + strata(etype), colon)
full.rnde.age = coxph(Surv(start, stop, status) ~ rx + nodes + differ + extent + age + strata(etype), colon)
full.rnde.obstruct = coxph(Surv(start, stop, status) ~ rx + nodes + differ + extent + obstruct + strata(etype), colon)
full.rnde.perfor = coxph(Surv(start, stop, status) ~ rx + nodes + differ + extent + perfor + strata(etype), colon)
full.rnde.adhere = coxph(Surv(start, stop, status) ~ rx + nodes + differ + extent + adhere + strata(etype), colon)
full.rnde.surg = coxph(Surv(start, stop, status) ~ rx + nodes + differ + extent + surg + strata(etype), colon)


AIC(full.rnde.sex, full.rnde.age, full.rnde.obstruct, full.rnde.perfor,
    full.rnde.adhere, full.rnde.surg)
```

Adding surg gives us the lowest AIC of 9311.77

```{r, include = FALSE}
full.rndes.sex = coxph(Surv(start, stop,status)~rx + nodes + differ+ extent + surg + sex + strata(etype), colon)
full.rndes.age = coxph(Surv(start, stop,status)~rx + nodes + differ+ extent + surg + age + strata(etype), colon)
full.rndes.perfor = coxph(Surv(start, stop,status)~rx + nodes + differ+ extent + surg + perfor + strata(etype), colon)
full.rndes.adhere = coxph(Surv(start, stop,status)~rx + nodes + differ+ extent + surg + adhere + strata(etype), colon)
full.rndes.obstruct = coxph(Surv(start, stop,status)~rx + nodes + differ+ extent + surg + obstruct + strata(etype), colon)


AIC(full.rndes.sex, full.rndes.age, full.rndes.perfor,
    full.rndes.adhere, full.rndes.obstruct)
```

Adding obstruct gives us an AIC of 9309.43
This change is minimal, so we will verify it using BIC.

```{r, include. = FALSE}
BIC(full.rnde.surg, full.rndes.obstruct)
```
By using BIC, which penalizes the number of covariates more strictly than AIC, we can see that the model with surg is a better model as it gives a BIC of 9340.13 as opposed to 9342.52 for the model that adds obstruct as another covariate. We will stop adding covariates here. 


```{r}
cox.zph(full.rnde.surg)

colon.fit.differ = survfit(Surv(start,stop, status) ~ differ, colon)
ggsurvplot(colon.fit.differ, colon, legend.labs=c('Well','Moderate', 'Poor'), fun='cloglog') + 
  labs(x='Days until Recurrence')

colon.fit.extent = survfit(Surv(start,stop, status) ~ extent, colon)
ggsurvplot(colon.fit.extent, colon, fun='cloglog') + 
  labs(x='Days until Recurrence')
```

Using a cox ZPH test and a log-log plot, we can see that extent and differ violate the PH assumption. To remedy this, we will construct a stratified model that stratifies on the 3 levels of tumor differentiation and 4 levels of local spread extent.   

```{r}

comb.strat <- coxph(Surv(start, stop, status) ~ rx + nodes + strata(differ) + strata(extent) + surg + strata(etype), data = colon)

cox.zph(comb.strat)
AIC(comb.strat)

```
We verify again with the cox ZPH test that none of our covariates are in violation of the proportional hazards assumption after stratifying on tumor differentiation. In addition, we have massively reduced our AIC score to 7188.06 fixing this violation. 

```{r}
exp(coef(comb.strat))
```

Using our model output, we can begin our model interpretation as follows:

-After accounting for other covariate effects, the impact on hazard rate of the Levamisole+5FU treatment is slightly increased to a factor of 0.854. In other words, its influence on decreasing hazard rate is slightly lessened after accounting for other variables. 
- For each additional lymph nodes with detectable cancer that is observed at the start of the study, the hazard rate for that individual increases by a factor of 1.064
-If time from surgery to registration is classified as 'long', hazard rate is increased by a factor of 1.16.


## Conclusion

Clearly, we have reached some interesting conclusions in our recurrence subset, death subset, and overall death & recurrence analysis. The covariate effects seem to be quite similar across all of these analyses and influence time until recurrence and death in similar ways. We can say with confidence that overall, low toxicity Levamisole is not an effective treatment in itself to treat colon cancer and reduce time until a recurrence or patient death. It must be combined with 5FU moderately toxic treatment to have a significant effect. The other measurements that were included in our model and discussed throughout our analyses are important indicators for researchers to predict which patients will have a higher rate of recurrence or death. Therefore, it is important when tracking patients to not only measure the treatment given, but the state of the disease itself at study onset. Measurements such as # nodes with cancer, extent of local spread, and time from surgery to registration are absolutely vital to make accurate predictions in addition to measuring the treatment given to a particular patient. If a patient has high local spread, many lymph nodes with detectable cancer, and a long time from surgery to registration, they can reasonably be classified as a 'high risk' for recurrence or death individual.  


\newpage
# 4. Citations
Laurie, J A et al. “Surgical adjuvant therapy of large-bowel carcinoma: an evaluation of levamisole and the combination of levamisole and fluorouracil. The North Central Cancer Treatment Group and the Mayo Clinic.” Journal of clinical oncology : official journal of the American Society of Clinical Oncology vol. 7,10 (1989): 1447-56. doi:10.1200/JCO.1989.7.10.1447

CG Moertel, TR Fleming, JS MacDonald, DG Haller, JA Laurie, CM Tangen, JS Ungerleider, WA Emerson, DC Tormey, JH Glick, MH Veeder and JA Maillard, Fluorouracil plus Levamisole as an effective adjuvant therapy after resection of stage II colon carcinoma: a final report. Annals of Internal Med, 122:321-326, 1991.

## R package citations

Therneau T (2024). _A Package for Survival Analysis in R_. R package
  version 3.8-3, <https://CRAN.R-project.org/package=survival>.

Terry M. Therneau, Patricia M. Grambsch (2000). _Modeling Survival
  Data: Extending the Cox Model_. Springer, New York. ISBN
  0-387-98784-3.

Kassambara A, Kosinski M, Biecek P (2024). _survminer: Drawing
  Survival Curves using 'ggplot2'_. R package version 0.5.0,
  <https://CRAN.R-project.org/package=survminer>.

H. Wickham. ggplot2: Elegant Graphics for Data Analysis.
  Springer-Verlag New York, 2016.

Wickham H, Vaughan D, Girlich M (2024). _tidyr: Tidy Messy Data_. R
  package version 1.3.1, <https://CRAN.R-project.org/package=tidyr>.


Xie Y (2025). _xfun: Supporting Functions for Packages Maintained by
  'Yihui Xie'_. R package version 0.52,
  <https://CRAN.R-project.org/package=xfun>.
