---
title: "PSTAT175-ProjectBrief"
author: "Kartik Adimulam"
date: "2025-04-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(survival)
library(survminer)



head(colon)
colon_fit = survfit(Surv(time, status) ~ rx, colon)

ggsurvplot(colon_fit, colon, title='KM Estimator Colon Cancer',xlab='time',
           ylab='Survival Prob.',surv.median.line = 'hv',palette = "Dark2", 
           conf.int = TRUE,    
           risk.table = TRUE,
           ggtheme = theme_minimal(), 
           font.main = c(16, "bold", "black"),  
           font.x = c(14, "bold"),            
           font.y = c(14, "bold"))
```

1. Our dataset contains survival time data of days until death of colon cancer patients that received different levels of chemotherapy. The primary covariate in our analysis is the 'rx' treatment variable which has 3 levels (No treatment, low-toxicity treatment, and moderate toxicity treatment) and the censoring status indicator. There are other covariates such as sex, age, obstruct(was the colon obstructed by the tumor?), perfor (was the colon perforated?), adhere (is the tumor on nearby organs?), nodes (how many lymph nodes have cancer?), differ (is the tumor differentiated?), extent (how much has the tumor spread locally?), surg (how long from surgery to registration?), node4 (indicator variable of more than 4 positive lymph nodes), and etype (did subject's cancer recur or death). The scientific questions we are trying to answer are: how do these covariates affect the probability of surival? are there some covariates that are not strong predictors and if so which ones? How much is survival probability affected by different covariate values? Do coviarates affect survival probability differently over time? Are there important interactions between these covariates? 

2. Laurie, J A et al. “Surgical adjuvant therapy of large-bowel carcinoma: an evaluation of levamisole and the combination of levamisole and fluorouracil. The North Central Cancer Treatment Group and the Mayo Clinic.” Journal of clinical oncology : official journal of the American Society of Clinical Oncology vol. 7,10 (1989): 1447-56. doi:10.1200/JCO.1989.7.10.1447

3. 